---
title: 'Lab1: RMarkdown'
author: "Lara Katz"
date: "2024-01-19"
output:
  word_document: default
  pdf_document: default
---

# Lab 1

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries:
```{r}
library(terra)
library(corrplot)
```


## Read in multivariate data set


```{r}
bds <- read.csv("./Data/Katz_BDS_data.csv", header = TRUE, row.names = 1) # Bridle shiner (BDS) data
bds <- bds[,-2] # remove 'ID' column (not informative, left over from random point generation)
```


## Use the head function to show the first few rows of the dataset


```{r}
head(bds)
```


## Bulleted list describing the columns of the dataset


Columns of dataset:

* **pab** = Presence/absence/pseudoabsence data (binary)

* **catchment**: Catchment position (stream size or size of largest input stream into lake/pond) (categorical data)

* **clay**: Mean clay composition (g/kg) of soil within point's HUC12 subwatershed (continuous data) 

* **coast**: Distance from the coastline in km (continuous data)

* **elev**: Mean elevation (m) within point's HUC12 subwatershed (continuous data)

* **lith**: Lithology at point (categorical data)

* **marine**: Whether or not point is within the historic marine limit (binary data)

* **ph**: Mean pH (*10) of soil within point's HUC12 subwatershed (continuous data) 

* Columns **X01_water** through **X29_AcAp_AlpineTundra**: Proportion of the LANDFIRE (https://www.landfire.gov/) Biophysical Settings landcover type within the point's HUC12 subwatershed (continuous data)
  + **X01_water**: Open Water
  + **X02_Barren**: Barren-Rock/Sand/Clay
  + **X03_LAc_NHardwd**: Laurentian-Acadian Northern Hardwoods Forest
  + **X05_NAtl_CoastPlain_Hardwd**: Northern Atlantic Coastal Plain Hardwood Forest
  + **X07_LAc_NPine.Oak**: Laurentian-Acadian Northern Pine(-Oak) Forest
  + **X08_LAc_Pine.Hemlock.Hardwd**: Laurentian-Acadian Pine-Hemlock-Hardwood Forest
  + **X09_CAp_Dry_Oak.Pine**: Central Appalachian Dry Oak-Pine Forest
  + **X10_Ap_Hemlock.N_Hardwood**: Appalachian (Hemlock-)Northern Hardwood Forest
  + **X11_Ac_Low.Elev_Spruce.Fir.Hardwood**: Acadian Low-Elevation Spruce-Fir-Hardwood Forest
  + **X12_AcAp_Montane_Spruce.Fir**: Acadian-Appalachian Montane Spruce-Fir Forest
  + **X13_CentAp_Pine.Oak_Rocky_Wd**:	Central Appalachian Pine-Oak Rocky Woodland
  + **X14_NAtl_Coast_Plain_Maritime**: Northern Atlantic Coastal Plain Maritime Forest
  + **X15_NAtl_Coast_Plain_Dune**: Northern Atlantic Coastal Plain Dune and Swale
  + **X16_CentIntAp_FloodplainSys**: Central Interior and Appalachian Floodplain Systems
  + **X17_CentIntAp_RiparianSys**: Central Interior and Appalachian Riparian Systems
  + **X18_LAc_FloodplainSys**: Laurentian-Acadian Floodplain Systems
  + **X19_Bor_Acidic_PeatSys**: Boreal Acidic Peatland Systems
  + **X20_CentIntAp_SwampSys**: Central Interior and Appalachian Swamp Systems
  + **X21_GulfAtl_CoastPlain_SwampSys**: Gulf and Atlantic Coastal Plain Swamp Systems
  + **X22_GulfAtl_CoastPlain_TMarshSys**:	Gulf and Atlantic Coastal Plain Tidal Marsh Systems	
  + **X23_LAc_Shrub.Herb_WetlSys**: Laurentian-Acadian Shrub-Herbaceous Wetland Systems
  + **X24_NCentInt_Wet_Flatwd**: North-Central Interior Wet Flatwoods
  + **X25_LAc_SwampSys**:	Laurentian-Acadian Swamp Systems
  + **X26_NeInt_PineBarrens**: Northeastern Interior Pine Barrens
  + **X27_AcAp_WdHeath.Krummholz**: Acadian-Appalachian Subalpine Woodland and Heath-Krummholz
  + **X28_Bor_JackPine.BlackSpruce**: Boreal Jack Pine-Black Spruce Forest
  + **X29_AcAp_AlpineTundra**: Acadian-Appalachian Alpine Tundra
* **sand**: Mean proportion sand composition (g/kg) of soil within point's HUC12 subwatershed (continuous data)

* **silt**: Mean proportion silt composition (g/kg) of soil within point's HUC12 subwatershed (continuous data)

* **slope**: Mean slope (degrees) within point's HUC12 subwatershed (continuous data)


## Plot data


Remove categorical data and rename variables:
```{r}
data.num <- as.data.frame(bds[,-c(1,2,6,7)]) # remove non-continuous data
data.num.names <- c("Clay",
                    "Coast",
                    "Elevation",
                    "pH",
                    "Water", 
                    "Barren", 
                    "LAc Northern Hardwoods", 
                    "NAtl Coast Plain Hardwood",
                    "LAc Northern Pine(-Oak)", 
                    "LAc Pine-Hemlock-Hardwood",
                    "CAp Dry Oak-Pine", 
                    "Ap (Hemlock-)N Hardwood",
                    "Ac Low-Elev Spruce-Fir-Hardwood",
                    "AcAp Montane Spruce-Fir",
                    "CAp Pine-Oak Rocky Wd", 
                    "NAtl Coast Plain Maritime",
                    "NAtl Coast Plain Dune",
                    "CentIntAp Floodplain Sys",
                    "CentIntAp Riparian Sys",
                    "LAc Floodplain Sys", 
                    "Bor Acidic Peat Sys", 
                    "CInt and Ap Swamp Sys", 
                    "Gulf/Atl Coast Plain Swamp Sys",
                    "Gulf/Atl Coast Pl,in TMarsh Sys",
                    "LAc Shrub-Herb Wetl Sys", 
                    "NCentInt Wet Flatwoods",
                    "LAc Swamp Sys",
                    "NInt Pine Barrens", 
                    "AcAp Wd/Heath-Krummholz",
                    "Bor Jack Pine-Black Spruce",
                    "AcAp Alpine Tundra", 
                    "Sand",
                    "Silt",
                    "Slope") # make land cover class names shorter
colnames(data.num) <- data.num.names

corr <- cor(data.num, method = "pearson") # Check variables for multicollinearity
# head(corr)
```

Plot correlogram:
```{r,fig.width=10,fig.height=10}
corrplot <- corrplot(corr, 
                     type = "lower",
                     method = "color",
                     diag = FALSE,
                     order = "FPC",
                     pch.col = "black",
                     pch.cex = 1,
                     tl.col = "black",
                     tl.cex = 1,
                     tl.srt = 45,
                     col = COL2(diverging = "RdYlBu"))
```

